---
title: "QC workflow - temperature and humidity"
author: "Patrick Kallabis"
date: "`r format(Sys.Date(), '%d.%m.%Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
---

This is a template. Fill in the SETTINGS block, then knit or run chunk by chunk. Every level is its own chunk with its parameters written out, so what ran is what you see. The defaults are the package defaults; the ones you will most likely touch are marked `# <-`.

Out of the box it runs on the Biel summer 2025 campaign that ships with the package (temperature only). That file is the clean raw campaign, so expect no flags - the version with planted errors is `Biel_test_tidy.csv` in tidy layout, see `tests/testthat/helper-testdata.R` for the pivot. Point the paths at your own files; the humidity part runs as soon as `rh_file` is set.

# Settings

## Load Packages

```{r setup, include = FALSE}

knitr::opts_chunk$set(message = TRUE, 
                      warning = TRUE, 
                      fig.width = 10, 
                      fig.height = 4)
```

```{r load all packages you might need }

load_packages <- function(cran = character(), github = character()) {
  # CRAN packages that are missing
  miss <- cran[!cran %in% rownames(installed.packages())]
  if (length(miss)) install.packages(miss, repos = "https://cloud.r-project.org")
  # GitHub packages as "owner/repo", installed through remotes
  for (g in github) {
    if (!basename(g) %in% rownames(installed.packages())) {
      if (!requireNamespace("remotes", quietly = TRUE))
        install.packages("remotes", repos = "https://cloud.r-project.org")
      remotes::install_github(g)
    }
  }
  invisible(lapply(c(cran, basename(github)), library, character.only = TRUE))
}

load_packages(cran   = c("xts", "zoo", "geosphere"),
              github = "Urban-Climate-Unibe/qcurs")
```

## Prepare for QC run

```{r settings}
# ---- files -------------------------------------------------------------
# temperature: wide table, one time column, one numeric column per logger
t_file   <- system.file("extdata", "Biel_summer_2025_raw.csv", package = "qcurs")   # <-
# humidity: same layout, same logger IDs; NULL if you have none
rh_file  <- NULL                                                                     # <-
# metadata: ID, LAT, LON, Landuse
md_file  <- system.file("extdata", "metadata.csv", package = "qcurs")                # <-
# where the results go
out_dir  <- "qc_output"                                                              # <-

# ---- how the files look --------------------------------------------------
csv_sep     <- ";"
csv_dec     <- "."
time_col    <- "time"                # name or number of the time column
time_format <- "%d.%m.%Y %H:%M"      # as in strptime()                             # <-
tz          <- "UTC"

# ---- which stations -------------------------------------------------------
# keep only the metadata rows of this city (NULL = all rows)
city        <- "BIEL"                                                                # <-

# ---- what to run ------------------------------------------------------------
run_humidity <- !is.null(rh_file)
interpolate  <- FALSE   # T9 / RH8 add values; leave off unless you need filled gaps

dir.create(out_dir, showWarnings = FALSE)
```

# Temperature

## 1. Read and check

The package does not read files. Read them here, look at them, and stop if the shape is wrong - a misread CSV is silent and everything downstream is wrong with it.

```{r t-read}
t_raw <- read.csv(t_file, sep = csv_sep, dec = csv_dec, check.names = FALSE,
                  fileEncoding = "UTF-8-BOM")
dim(t_raw)
head(t_raw[, 1:min(6, ncol(t_raw))])
# every logger column must be numeric; a decimal comma or a text column shows up here
str(t_raw[, 1:min(6, ncol(t_raw))])
```

```{r t-convert}
# xts with one column per logger; prints a conversion report
t_x <- qc_as_xts(t_raw, time_col = time_col, time_format = time_format, tz = tz)
```

## 2. Metadata

```{r metadata}
md <- read.csv(md_file, sep = csv_sep, dec = csv_dec, stringsAsFactors = FALSE,
               fileEncoding = "UTF-8-BOM")
if (!is.null(city)) md <- md[md$City == city, ]
stopifnot(all(c("ID", "LAT", "LON", "Landuse") %in% names(md)))
table(md$Landuse)
# loggers in the data without a metadata row cannot be judged by levels 6 and 7
setdiff(colnames(t_x), md$ID)
```

## 3. The chain, level by level

Every level takes the list the previous one returned and gives it back with its flags and its record added. Change a parameter, re-run from that chunk on.

```{r t1}
t_res <- T_QC_1_gross_error(t_x,
                            threshold_max = 60,
                            threshold_min = -40)
```

```{r t2}
# seasonal bounds; summer_min is the one worth checking for cold rural sites
t_res <- T_QC_2_out_of_range(t_res,
                             summer_min = -5.85)                                      # <-
```

```{r t3}
t_res <- T_QC_3_time_consistency(t_res,
                                 dt = 3,          # half window in steps
                                 threshold = 6)   # K
```

```{r t4}
# the window is also the detection delay in live data; check winter plateaus
# before shortening it
t_res <- T_QC_4_stuck_values(t_res,
                             window_size = "6 hours",                                 # <-
                             na_tolerance_frac = 0.5,
                             min_non_na = 5,
                             sd_tol = 0)
```

```{r t5}
# refuses on a single-season campaign; set require_multi_year = FALSE only if
# you know what a one-year monthly IQR means for your data
t_res <- T_QC_5_climatic_outliers(t_res,
                                  ext_lim_factor = 4,
                                  min_n = 300,
                                  require_multi_year = TRUE)
```

```{r t6}
t_res <- T_QC_6_spatial_consistency(t_res, metadata = md,
                                    k_sigma = 6,
                                    abs_floor = 3,        # K
                                    radius_m = 3000,
                                    k_neighbours = 5,
                                    min_neighbours = 2,
                                    landuse_mode = "graded")
```

```{r t7}
t_res <- T_QC_7_spatiotemporal_consistency(t_res, metadata = md,
                                           radius_m = 2500,
                                           n_neighbours = 5,
                                           threshold = 0.99)
```

```{r t8}
t_res <- T_QC_8_diurnal_range(t_res,
                              range_frac = 0.4,
                              min_consecutive_days = 2,
                              min_obs_per_day = 72,
                              min_reference_days = 14)
```

```{r t9, eval = interpolate}
# ADDS values. Codes 50 (plain gap) and 50 + level (removed, then refilled)
t_res <- T_QC_9_interpolate(t_res,
                            maxgap = 5,
                            refill_flagged = TRUE)
```

## 4. What came out

```{r qc-summary-fun, include = FALSE}
# one row per level and station: flagged, judged, and judged as share of the values
qc_summary <- function(res, chain = c("t", "rh")) {
  chain <- match.arg(chain)
  keys  <- grep(paste0("^", chain, "[0-9]_"), names(res$qc_info), value = TRUE)
  rows  <- lapply(keys, function(k) {
    r <- res$qc_info[[k]]
    if (is.null(r$n_flagged_by_station)) return(NULL)   # skipped levels, interpolation
    data.frame(level = k, station = names(r$n_flagged_by_station),
               flagged = as.integer(r$n_flagged_by_station),
               judged  = as.integer(r$n_judged_by_station), stringsAsFactors = FALSE)
  })
  out <- do.call(rbind, rows)
  n_valid <- colSums(!is.na(coredata(res$qc_data)) | coredata(res$qc_data_flagged) > 0)
  out$judged_pct <- round(100 * out$judged / n_valid[out$station], 1)
  out
}
```

```{r t-summary}
FLG <- coredata(t_res$qc_data_flagged)
# how many cells carry which code
table(FLG, useNA = "ifany")
# per level and station: what was flagged, and what the level actually looked at
t_sum <- qc_summary(t_res, "t")
# flagged
xtabs(flagged ~ station + level, t_sum)
# judged, as percent of the station's values - 0 means the level never reached it
xtabs(judged_pct ~ station + level, t_sum)
```

Where a level was blind, by name:

```{r t-refusals}
t_res$qc_info$t5_climatic_outliers$refused
t_res$qc_info$t6_spatial_consistency$never_evaluated
t_res$qc_info$t7_spatiotemporal_consistency$skipped
t_res$qc_info$t8_diurnal_range$skipped_stations
```

One station, raw against cleaned, flagged points coloured by level:

```{r t-plot}
station <- colnames(t_x)[1]                                                          # <-
raw <- t_x[, station]; cln <- t_res$qc_data[, station]; f <- t_res$qc_data_flagged[, station]
plot(index(raw), coredata(raw), type = "l", col = "grey70", xlab = "", ylab = "deg C",
     main = station)
lines(index(cln), coredata(cln), col = "black")
codes <- sort(unique(coredata(f)[coredata(f) > 0 & !is.na(coredata(f))]))
pal <- c("1" = "red", "2" = "orange", "3" = "purple", "4" = "blue", "5" = "brown",
         "6" = "darkgreen", "7" = "magenta", "8" = "cyan4")
for (cc in codes[codes < 50])
  points(index(raw)[coredata(f) == cc], coredata(raw)[coredata(f) == cc], col = pal[as.character(cc)], pch = 16, cex = 0.6)
if (length(codes[codes < 50]))
  legend("topleft", legend = paste("level", codes[codes < 50]), col = pal[as.character(codes[codes < 50])],
         pch = 16, bty = "n", ncol = 4)
```

## 5. Export

Cleaned series, flags and the record, side by side. The flag file is the audit trail; keep it with the data.

```{r t-export}
as_table <- function(x) data.frame(time = format(index(x), "%Y-%m-%d %H:%M:%S"), coredata(x),
                                   check.names = FALSE)
write.csv(as_table(t_res$qc_data),         file.path(out_dir, "temperature_qc.csv"),       row.names = FALSE)
write.csv(as_table(t_res$qc_data_flagged), file.path(out_dir, "temperature_flags.csv"),    row.names = FALSE)
write.csv(t_sum,                            file.path(out_dir, "temperature_summary.csv"),  row.names = FALSE)
saveRDS(t_res, file.path(out_dir, "temperature_result.rds"))   # the whole thing, qc_info included
capture.output(str(t_res$qc_info, max.level = 2, give.attr = FALSE),
               file = file.path(out_dir, "temperature_qc_info.txt"))
list.files(out_dir)
```

# Humidity

Same logger IDs, same time stamps as the temperature. Cells without a temperature twin at exactly the same stamp are not judged by levels 1, 5 and 7.

```{r rh-read, eval = run_humidity}
rh_raw <- read.csv(rh_file, sep = csv_sep, dec = csv_dec, check.names = FALSE,
                   fileEncoding = "UTF-8-BOM")
dim(rh_raw)
rh_x <- qc_as_xts(rh_raw, time_col = time_col, time_format = time_format, tz = tz)
# loggers with humidity but no temperature
setdiff(colnames(rh_x), colnames(t_x))
```

```{r rh1, eval = run_humidity}
# pass the WHOLE temperature result: the flags are inherited from it
rh_res <- RH_QC_1_inherit_temperature(rh_x, temperature_flags = t_res,
                                      inherit_gapfill = FALSE)
```

```{r rh2, eval = run_humidity}
# up to 105 is kept: condensation tolerance of the element
rh_res <- RH_QC_2_range(rh_res,
                        rh_min = 0,
                        rh_max = 105)
```

```{r rh3, eval = run_humidity}
rh_res <- RH_QC_3_spike(rh_res,
                        dt = 3,
                        threshold = 20)   # percent RH
```

```{r rh4, eval = run_humidity}
rh_res <- RH_QC_4_stuck_values(rh_res,
                               window_size = "6 hours",                               # <-
                               sd_tol = 0.1,
                               sat_max = 95,          # constant at saturation is not a fault
                               min_valid_frac = 0.5)
```

```{r rh5, eval = run_humidity}
# uses the CLEANED temperature of t_res
rh_res <- RH_QC_5_decoupling(rh_res, temperature = t_res,
                             decouple_r = -0.3,
                             decouple_trange = 3,
                             decouple_min_n = 60)
```

```{r rh6, eval = run_humidity}
# a drifter stays below station_max while the network median sits above network_min
rh_res <- RH_QC_6_saturation_drift(rh_res,
                                   network_min = 95,
                                   station_max = 90,
                                   min_hits = 12,
                                   min_stations = 3,
                                   window_days = 30)                                  # <-
```

```{r rh7, eval = run_humidity}
rh_res <- RH_QC_7_dewpoint_consistency(rh_res, temperature = t_res, metadata = md,
                                       k_sigma = 6,
                                       abs_floor = 3,
                                       radius_m = 3000,
                                       k_neighbours = 5,
                                       min_neighbours = 2,
                                       landuse_mode = "graded")
```

```{r rh8, eval = run_humidity && interpolate}
rh_res <- RH_QC_8_interpolate(rh_res,
                              maxgap = 5,
                              refill_flagged = TRUE)
```

```{r rh-summary, eval = run_humidity}
table(coredata(rh_res$qc_data_flagged), useNA = "ifany")
rh_sum <- qc_summary(rh_res, "rh")
xtabs(flagged ~ station + level, rh_sum)
xtabs(judged_pct ~ station + level, rh_sum)
# which temperature levels the humidity inherited from (the RH code is always 1)
rh_res$qc_info$rh1_inherit_temperature$n_flagged_by_t_code
rh_res$qc_info$rh5_decoupling$skipped_stations
rh_res$qc_info$rh7_dewpoint_consistency$never_evaluated
```

```{r rh-export, eval = run_humidity}
write.csv(as_table(rh_res$qc_data),         file.path(out_dir, "humidity_qc.csv"),      row.names = FALSE)
write.csv(as_table(rh_res$qc_data_flagged), file.path(out_dir, "humidity_flags.csv"),   row.names = FALSE)
write.csv(rh_sum,                            file.path(out_dir, "humidity_summary.csv"), row.names = FALSE)
saveRDS(rh_res, file.path(out_dir, "humidity_result.rds"))
```

# Flag codes

| code | temperature | humidity |
|----|----|----|
| `NA` | no observation | no observation |
| `0` | not objected by any level that reached the cell | same |
| `1` | gross error | inherited from temperature |
| `2` | seasonal range | range 0-105 |
| `3` | spike | spike |
| `4` | stuck | stuck |
| `5` | climatic outlier | decoupled from own T |
| `6` | spatial consensus | saturation drift |
| `7` | spatiotemporal | dewpoint consensus |
| `8` | diurnal range collapse | \- |
| `50` | plain gap filled | same |
| `50+N` | removed by N, refilled | same |

```{r session}
sessionInfo()
```
